---
title: "PMConnolly Package"
author: "Thomas Connolly"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PMConnolly Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The PMConnolly package was created to implement the statistical procedures described in the paper written by Pei Fen Kuan and Bo Huang^[A Simple and Robust Method for Partially Matched Samples Using the P-Values Pooling Approach](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3717400/). When collecting data the problem of missing observations is unfortunately a very common issue. Partially matched data is when two datasets are said to be matched pairs but there are observations missing from one or both said datasets. The paper discusses five methods for dealing with partially matched samples and uses the example of tumor and corresponding normal tissue samples as its two partially matched datasets. Both datasets have missing values and the procedures described are then applied on the data to show their effectiveness.

The procedures included in the paper and within this package include:

1. Liptak's Weighted Z-test
1. Kim et al's Modified t-statistic
1. Looney and Jones's Corrected Z-test
1. Lin and Stivers's MLE Based Test Under Heteroscedasticity
1. Ekbohm's MLE Based Test Under Homoscedasticity

### Setup
```{r setup}
library(PMConnolly)
```
### Liptak's Weighted Z-test

#### Usage
```{r eval=FALSE}
liptak_weighted(x,y,alternative="two.sided")
```

#### Arguments
* $x$ a (non-empty) numeric vector of data values
* $y$ a (non-empty) numeric vector of data values
* *alternative* a character string specifying the alternative hypothesis, must be one of the following: "two.sided"(default), "greater", or "less"

#### Formulas 
The combined p-value is caclulated by:
$$
\begin{aligned}
 p_{ci}=1-\Phi(\frac{w_1Z_{1i}+w_2Z_{2i}}{\sqrt{w_1^2+w_2^2}})
\end{aligned}
$$
where
$$
\begin{aligned}
z_{ai}=\Phi^{-1}(1-p_{ai})
,a=1,2
\end{aligned}
\\
\begin{aligned}
w_1=\sqrt{2n_1}
\end{aligned}
\\
\begin{aligned}
w_2=\sqrt{n_2+n_3}
\end{aligned}
$$
For a two sided test the p-value is:
$$
\begin{aligned}
p_{ci}^*=\begin{cases} 2p_{ci} & p_{ci}<1/2\\
2(1-p_{ci}) & otherwise
\end{cases}
\end{aligned}
$$

#### Examples
```{r}
x <- c(150,210,NA,350,315,NA,NA,98,170,165)
y <- c(NA,150,305,180,NA,90,80,206,111,43)
liptak_weighted(x,y,alternative="two.sided")
liptak_weighted(x,y,alternative="greater")
liptak_weighted(x,y,alternative="less")
```


### Kim et al's Modified t-statistic

#### Usage
```{r eval = F}
kim_mod(x,y,alternative="two.sided")
```

#### Arguments
* $x$ a (non-empty) numeric vector of data values
* $y$ a (non-empty) numeric vector of data values
* *alternative* a character string specifying the alternative hypothesis, must be one of the following: "two.sided"(default), "greater", or "less"

#### Formulas
The modified t-statistic is:
$$
\begin{aligned}
 t_3=\frac{{n_1}\bar{D}+{n_H}(\bar{T}-\bar{N})}{\sqrt{{n_1}{S_D}^2+{n_H}^2(S_N^2/n_3+S_T^2/n_2)}}
\end{aligned}
$$

#### Example
```{r}
x <- c(150,210,NA,350,315,NA,NA,98,170,165)
y <- c(NA,150,305,180,NA,90,80,206,111,43)
kim_mod(x,y,alternative="two.sided")
kim_mod(x,y,alternative="greater")
kim_mod(x,y,alternative="less")
```


### Looney and Jones's Corrected Z-test

#### Usage
```{r eval = F}
looney_jones(x,y,alternative="two.sided")
```

#### Arguments
* $x$ a (non-empty) numeric vector of data values
* $y$ a (non-empty) numeric vector of data values
* *alternative* a character string specifying the alternative hypothesis, must be one of the following: "two.sided"(default), "greater", or "less"

#### Formulas
The corrected Z-statistic is given by:
$$
\begin{aligned}
 Z_{corr}=\frac{\bar{T^*}-\bar{N^*}}{\sqrt{S_T^{*2}/(n_1+n_2)+S_N^{*2}/(n_1+n_3)-2n_1S_{TN_1}/(n_1+n_2)(n_1+n_3)}}
\end{aligned}
$$

#### Examples
```{r}
x <- c(150,210,NA,350,315,NA,NA,98,170,165)
y <- c(NA,150,305,180,NA,90,80,206,111,43)
looney_jones(x,y,alternative="two.sided")
looney_jones(x,y,alternative="greater")
looney_jones(x,y,alternative="less")
```


### Lin and Stivers's MLE Based Test Under Heteroscedasticity

#### Usage
```{r eval = F}
lin_stivers(x,y,alternative="two.sided")
```

#### Arguments
* $x$ a (non-empty) numeric vector of data values
* $y$ a (non-empty) numeric vector of data values
* *alternative* a character string specifying the alternative hypothesis, must be one of the following: "two.sided"(default), "greater", or "less"

#### Formulas
The MLE based test statistic is given by:
$$
\begin{aligned}
 Z_{LS}=\frac{f(\bar{T_1}-\bar{T})-g(\bar{N_1}-\bar{N})+\bar{T}-\bar{N}}{\sqrt{V_1}}
\end{aligned}
$$
where
$$
\begin{aligned}
V_{1}=\frac{(f^{2}/n_{1}+(1-f)^{2}/n_{2})S^{2}_{T_{1}}(n_{1}-1)+(g^{2}/n_{1}+(1-g)^{2}/n_{3})S^{2}_{N_{1}}(n_{1}-1)-2fgS_{TN_{1}}(n_{1}-1)/n1}{(n_{1}-1)}
\end{aligned}
\\
\begin{aligned}
f=n_{1}(n_{1}+n_{3}+n_{2}S_{TN_{1}}/S^{2}_{T_{1}})((n_{1}+n_{2})(n_{1}+n_{3})-n_{2}n_{3}r^{2})^{-1}
\end{aligned}
\\
\begin{aligned}
g=n_{1}(n_{1}+n_{2}+n_{3}S_{TN_{1}}/S^{2}_{N_{1}})((n_{1}+n_{2})(n_{1}+n_{3})-n_{2}n_{3}r^{2})^{-1}
\end{aligned}
\\
\begin{aligned}
r=\frac{S_{TN_{1}}}{S_{T_{1}}S_{N_{1}}}
\end{aligned}
$$

#### Examples
```{r}
x <- c(150,210,NA,350,315,NA,NA,98,170,165)
y <- c(NA,150,305,180,NA,90,80,206,111,43)
lin_stivers(x,y,alternative="two.sided") #Z_LS=1.584755 p_value=0.1738763 df=5
lin_stivers(x,y,alternative="greater") #Z_LS=1.584755  p_value=0.08693813 df=5
lin_stivers(x,y,alternative="less") #Z_LS=1.584755 p_value=0.9130619 df=5
```


### Ekbohm's MLE Based Test Under Homoscedasticity

#### Usage
```{r eval = F}
ekbohm(x,y,alternative="two.sided")
```

#### Arguments
* $x$ a (non-empty) numeric vector of data values
* $y$ a (non-empty) numeric vector of data values
* *alternative* a character string specifying the alternative hypothesis, must be one of the following: "two.sided"(default), "greater", or "less"

#### Formulas
The MLE based test statistic is given by:
$$
\begin{aligned}
 Z_{E}=\frac{f^{*}(\bar{T_1}-\bar{T})-g^{*}(\bar{N_1}-\bar{N})+\bar{T}-\bar{N}}{\sqrt{V^{*}_1}}
\end{aligned}
$$
where
$$
\begin{aligned}
V^{*}_{1}=\hat{\sigma}^{2}\frac{2n_{1}(1-r)+(n_{2}+n_{3})(1-r^{2})}{(n_{1}+n_{2})(n_{1}+n_{3})-n_{2}n_{3}r^{2}}
\end{aligned}
\\
\begin{aligned}
\hat{\sigma}^{2}=\frac{S^{2}_{T_{1}}(n_{1}-1)+S^{2}_{N_{1}}(N_{1}-1)+(1+r^{2})(S^{2}_{T}(n_{2}-1)+S^{2}_{N}(n_{3}-1))} {2(n_{1}-1)+(1+r^{2})(n_{2}+n_{3}-2)}
\end{aligned}
\\
\begin{aligned}
f^{*}=n_{1}(n_{1}+n_{3}+n_{2}r)((n_{1}+n_{2})(n_{1}+n_{3})-n_{2}n_{3}r^{2})^{-1}
\end{aligned}
\\
\begin{aligned}
g^{*}=n_{1}(n_{1}+n_{2}+n_{3}r)((n_{1}+n_{2})(n_{1}+n_{3})-n_{2}n_{3}r^{2})^{-1}
\end{aligned}
\\
\begin{aligned}
r=\frac{S_{TN_{1}}}{S_{T_{1}}S_{N_{1}}}
\end{aligned}
$$

#### Examples
```{r}
x <- c(150,210,NA,350,315,NA,NA,98,170,165)
y <- c(NA,150,305,180,NA,90,80,206,111,43)
ekbohm(x,y,alternative="two.sided") #Z_E=1.355812 p_value=0.2331778 df=5
ekbohm(x,y,alternative="greater") #Z_E=1.355812  p_value=0.1165889 df=5
ekbohm(x,y,alternative="less") #Z_E=1.355812 p_value=0.8834111 df=5
```

